---
title: "Practical 3: Parameter Estimation — Solutions"
format:
  html:
    toc: true
execute:
  warning: false
---

> Posted after the tutorial group. Try the exercises yourself first — the value is in getting stuck, not in
> reading the answer.

## Setup (from the handout)

```{r}
intensity <- 1:9
n_trials <- rep(40, 9)
n_yes <- c(1, 4, 5, 11, 19, 24, 35, 38, 39)
prop_yes <- n_yes / n_trials

data <- data.frame(intensity, n_yes, n_trials, prop_yes)

psychometric_model <- function(mu, sigma, x) {
  pnorm(x, mean = mu, sd = sigma)
}

rmsd <- function(params, data) {
  mu <- params[1]; sigma <- params[2]
  predicted <- psychometric_model(mu, sigma, data$intensity)
  sqrt(mean((data$prop_yes - predicted)^2))
}
```

## Question 1

There's no single correct numeric answer here — it's an eyeballing exercise. A reasonable guess from looking
at the plot: `mu` somewhere around 4–6 (wherever the curve visually crosses 50%), and `sigma` somewhere
around 0.5–1.5 (a moderately gradual, not extremely sharp or shallow, transition). This is just a benchmark
to compare against later — don't worry about whether your guess was "right."

## Question 2

```{r}
rmsd(c(5, 1), data)     # a plausible guess
rmsd(c(2, 3), data)     # a clearly worse guess
rmsd(c(5, 0.9), data)   # another plausible guess
```

You should find that guesses closer to `mu ≈ 5, sigma ≈ 1` give a lower RMSD. Try your own Question-1 guess
in the same way and see how it scores.

## Question 3

```{r}
mu_vals <- seq(1, 9, length.out = 100)
sigma_vals <- seq(0.1, 3, length.out = 100)
grid <- expand.grid(mu = mu_vals, sigma = sigma_vals)
grid$rmsd <- apply(grid, 1, function(row) rmsd(row, data))
best <- grid[which.min(grid$rmsd), ]
best
```

Your eyeballed guess was probably in the right neighborhood but not exact — grid search finds a combination
close to `mu ≈ 5, sigma ≈ 1`. The dark "valley" in the `levelplot` is the region of `(mu, sigma)`
combinations that all produce a similarly low RMSD — i.e., curves that are all roughly compatible with the
data. It's worth noticing that this valley has some width to it, not just a single point — that's a preview
of identifiability issues you'll meet again with more complex models.

## Question 4

```{r}
fit_extreme <- optim(par = c(mu = 1, sigma = 2.9), fn = rmsd, data = data)
fit_extreme$par
```

You should find `optim()` still recovers essentially the same answer as before, even from a poor starting
point — the RMSD surface for this simple, well-behaved two-parameter problem has just one clear valley (you
can see this in the `levelplot`), so there's nowhere else for Nelder-Mead to get stuck. This connects back to
the lecture's discussion of local minima: this particular landscape happens to be well-behaved, but that's
not guaranteed for more complex models, which is why running `optim()` from several different starting
values is a good habit to build now.

## Question 5

```{r}
fit_rmsd <- optim(par = c(mu = 4, sigma = 0.5), fn = rmsd, data = data)
fit_rmsd$par
```

You should trust `fit_rmsd$par` more than your Question 1 eyeball guess — `optim()` is systematically
minimizing RMSD, while eyeballing is just a visual impression. There's genuinely no "true" value to compare
either one to here, which is worth sitting with for a moment: most real data you'll work with in your own
research has no known ground truth either, so a systematic, repeatable fitting procedure matters even more
than it might seem to from today's earlier examples, where you sometimes do get to check against a known
answer.

## Question 6

```{r}
nll <- function(params, data) {
  mu <- params[1]; sigma <- params[2]
  predicted <- psychometric_model(mu, sigma, data$intensity)
  predicted <- pmin(pmax(predicted, 1e-6), 1 - 1e-6)
  -sum(dbinom(data$n_yes, data$n_trials, predicted, log = TRUE))
}

fit_mle <- optim(par = c(mu = 4, sigma = 0.5), fn = nll, data = data)
fit_mle$par
fit_rmsd$par
```

The two fits should be very close to each other. Neither is "more correct" in any general sense for this
dataset — for binomial count data like this, MLE is the statistically principled choice, but RMSD happens to
land in a similar place here because the data is fairly clean.

## Question 7 (discussion)

Here's a concrete way to see the difference. Compare two situations where the model misses the observed
count by the same amount — 2 "yes" responses out of 40:

```{r}
# a miss of 2, near the middle of the range (prop_yes close to 0.5)
dbinom(18, size = 40, prob = 0.45)   # observed 18, model predicted 0.45 -> "expected" ~18
dbinom(20, size = 40, prob = 0.45)   # observed 20 instead - a miss of 2

# a miss of 2, near the edge of the range (prop_yes close to 0)
dbinom(0, size = 40, prob = 0.05)    # observed 0, model predicted 0.05 -> "expected" ~2
dbinom(2, size = 40, prob = 0.05)    # observed 2 instead - the same miss of 2
```

Run this yourself and compare the probabilities. RMSD scores both misses identically, since it only looks at
the size of the gap (2, in both cases) and doesn't care where on the curve it happens. Likelihood behaves
differently: a miss of a given size near the edges of the range (close to 0 or 1) typically changes the
probability more dramatically than the same-sized miss near the middle (close to 0.5), because the binomial
distribution is naturally more "spread out" in the middle and more "compressed" near its boundaries. That's
what "counts near the boundaries" means in practice — an observed count like 0 or 40 out of 40, as opposed to
something like 18 or 20.

The "noise doesn't match the assumptions" part is about a different scenario: `dbinom()` assumes each of the
40 trials at a given intensity is independent with a fixed underlying probability. If that assumption is
wrong — for example, if a participant's responses drift over the course of the block, or trials aren't truly
independent (maybe they tend to repeat their previous answer) — the *actual* variability in the data won't
match what the binomial model expects, and MLE's built-in assumptions about likely vs. unlikely data will be
calibrated to the wrong thing. RMSD makes no such assumption at all — it just measures distance — so it's
less sensitive to that particular kind of mismatch, at the cost of not having the same statistical guarantees
MLE has when its assumptions do hold.

## Part 4: Simulating your own data

```{r}
set.seed(99)
intensity_sim <- 1:9
true_mu <- 4
true_sigma <- 1.3
p_true <- pnorm(intensity_sim, mean = true_mu, sd = true_sigma)

n_trials_sim <- rep(40, 9)
n_yes_sim <- rbinom(length(intensity_sim), n_trials_sim, p_true)
prop_yes_sim <- n_yes_sim / n_trials_sim

data_sim <- data.frame(intensity = intensity_sim, n_yes = n_yes_sim,
                        n_trials = n_trials_sim, prop_yes = prop_yes_sim)
```

## Question 8

No single correct answer — an eyeballing exercise, same as Question 1.

## Question 9

```{r}
fit_sim <- optim(par = c(mu = 4, sigma = 0.5), fn = rmsd, data = data_sim)
fit_sim$par
```

You should get an estimate somewhere close to `mu ≈ 4, sigma ≈ 1.3` (exact values will vary a little
depending on your random seed and starting values, which is fine).

## Question 10

```{r}
c(true_mu = true_mu, true_sigma = true_sigma)
fit_sim$par
```

Your fit should sit close to the true values — a satisfying confirmation that the fitting method actually
recovers the parameters that generated the data, when there's noise involved but the underlying model is
correct. Your Question 8 eyeball guess was probably in the right neighborhood but visibly less accurate than
`optim()`'s fit — the same lesson as Question 5, just now with a ground truth to check against directly.

## Question 11

```{r}
plot(data_sim$intensity, data_sim$prop_yes, pch = 19,
     xlab = "Stimulus intensity", ylab = "Proportion 'yes' responses", ylim = c(0, 1))

x_seq <- seq(1, 9, length.out = 200)
lines(x_seq, psychometric_model(fit_sim$par[1], fit_sim$par[2], x_seq), col = "blue")
```

If your line didn't show up in RStudio, the most common causes are: the `lines()` call still had a `#` in
front of it (comment it back in), one of the two blanks in the scaffold wasn't replaced with an actual value,
or the plot and the `lines()` call ended up in separate chunks that were run out of order — a `lines()` call
needs an existing `plot()` to draw onto, so re-run them together in one go if this happens. Once it works,
the fitted curve should pass closely through your observed points.

## Question 12 (interpretation)

Comparing `mu ≈ 4` for this simulated participant to `mu ≈ 5` for the Part 0–3 participant: this participant
has a **lower threshold**, meaning they detect fainter stimuli — they're more sensitive overall. Their
estimated `sigma` (≈1.3) is a bit larger than the Part 0–3 participant's (≈1), meaning their transition is
slightly more gradual — somewhat less reliable/precise responding around threshold. The two parameters have
separate, interpretable psychological meanings (overall sensitivity vs. reliability of responding) — worth
keeping in mind, since this distinction matters again when comparing parameters across conditions or groups
in later sessions.

## Part 5: Does the shape of the model matter?

**Question 13.**

```{r}
linear_model <- function(a, b, x) {
  a + b * x
}
```

**Question 14.**

```{r}
rmsd_linear <- function(params, data) {
  a <- params[1]; b <- params[2]
  predicted <- linear_model(a, b, data$intensity)
  sqrt(mean((data$prop_yes - predicted)^2))
}
```

**Question 15.**

```{r}
fit_linear <- optim(par = c(a = 0, b = 0.1), fn = rmsd_linear, data = data)
fit_linear$par
fit_linear$value
```

Your fitted slope `b` should come out positive, since "yes" responses increase with intensity — a good sanity
check that your fit is behaving sensibly.

**Question 16.**

```{r}
fit_linear$value
fit_rmsd$value
```

On this dataset, the cumulative normal model's RMSD is typically somewhat lower than the linear model's,
though the difference may be less dramatic than you'd expect over the *middle* of the intensity range, where
a straight line can approximate an S-curve reasonably well. The real story shows up visually in Question 17,
not just in the single RMSD number.

**Question 17.**

```{r}
plot(data$intensity, data$prop_yes, pch = 19, ylim = c(-0.2, 1.2),
     xlab = "Intensity", ylab = "Proportion 'yes' responses")

x_seq <- seq(1, 9, length.out = 200)
lines(x_seq, psychometric_model(fit_rmsd$par[1], fit_rmsd$par[2], x_seq), col = "darkgreen")
lines(x_seq, linear_model(fit_linear$par[1], fit_linear$par[2], x_seq), col = "blue")
abline(h = c(0, 1), lty = "dotted")
legend("topleft", legend = c("Cumulative normal fit", "Linear fit"),
       col = c("darkgreen", "blue"), lty = 1)
```

(Note: the y-axis is deliberately extended a bit beyond 0–1 here so the linear model's out-of-bounds
predictions are visible — widen your own `ylim` if you don't see this.) The two curves typically agree
reasonably well through the middle intensities, but diverge clearly at the low and high ends.

**Question 18 (discussion).** Check the linear model's prediction at the extremes: at `intensity = 1`, the
linear fit will typically predict a proportion *below* 0 (a negative "probability" of responding yes), and at
`intensity = 9` it will often predict a proportion *above* 1. Both are impossible for something that has to
be a proportion between 0 and 1. The cumulative normal model, by construction, can never do this — it
smoothly approaches (but never crosses) 0 and 1 at the extremes, which is exactly why S-shaped functions are
the standard choice for this kind of bounded data. This is a good example of a model being flexible enough to
track the *observed* data reasonably well while still making an implausible prediction outside that
comfortable middle range — precisely the kind of issue formal model comparison methods (Weeks 6–7) are built
to catch systematically, rather than relying on eyeballing a plot.
