---
title: "Parameter Estimation"
subtitle: "Basic Methods & Maximum Likelihood"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    code-line-numbers: true
    scrollable: true
execute:
  echo: true
  warning: false
  fig-width: 8
  fig-height: 5.5
---

## Where we are

- Last week: what *is* a model, and why do psychologists build them?
- Today: once you have a model, how do you find the numbers that make it fit your data?

---

## Learning objectives

By the end of today you should be able to:

1. Explain what a parameter is, and why we need to *estimate* it from data
2. Explain, in words, what the scale and shape parameters of a Weibull psychometric function represent
3. Explain what RMSD is, and why data never sits exactly on a fitted curve
4. Use grid search to find best-fitting parameters by hand, and explain why it doesn't scale
5. Explain, at a high level, how `optim()`'s default method (Nelder-Mead simplex) searches — and its
   limitations, including local minima
6. Explain the idea of maximum likelihood, and how it relates to simulating data
7. Fit a model using both RMSD and maximum likelihood

---

## You already know a model like this

$$y = a + b \cdot x$$

- A straight line has two parameters too: an intercept (`a`) and a slope (`b`)
- Fitting a line to data — the way you may have done in a stats course — is exactly the same problem we're
  about to solve today, just with a different-shaped curve
- We won't fit a line today, but keep it in mind — the tutorial will ask you to

---

## Luminance spot detection

```{r}
#| echo: false
set.seed(42)
n_spots <- 18
spot_x <- runif(n_spots, 0.5, 9.5)
spot_y <- runif(n_spots, 0.3, 1.7)
spot_brightness <- runif(n_spots, 0.05, 1)

old_par <- par(bg = "black", mar = c(2, 1, 3, 1))
plot(1, type = "n", xlim = c(0, 10), ylim = c(0, 2), axes = FALSE, xlab = "", ylab = "")
points(spot_x, spot_y, pch = 19, cex = 4, col = gray(spot_brightness))
par(old_par)
```

- Some spots barely differ from the black background — participants will often say "no" even when one was
  actually shown
- Some are unmistakably brighter than the background — participants will almost always say "yes"
- In a real experiment, we control the exact luminance of each spot systematically, and present many trials
  at each level — next, one trial up close

::: {.notes}
This is a schematic illustration, not photographs from an actual experiment — the point is to make the
detection task concrete before the data table appears. Point out explicitly that the "interesting" middle
spots are exactly where the psychometric function has its steepest slope.
:::

---

## A concrete example: one trial, up close

```{r}
#| echo: false
old_par <- par(bg = "black", mar = c(2, 1, 2, 1))
plot(1, type = "n", xlim = c(0, 2), ylim = c(0, 2), axes = FALSE, xlab = "", ylab = "")
points(1, 1.3, pch = 19, cex = 5, col = gray(0.22))
text(1, 0.55, "Do you see the dot?", col = "white", cex = 1.6)
text(1, 0.2, "Yes  /  No", col = "white", cex = 1.8, font = 2)
par(old_par)
```

- This is literally what one trial looks like from the participant's side: one stimulus, one binary answer
- Repeat this many times per intensity, and the *proportion* of "yes" answers is what traces out the curve
  on the previous slide

::: {.notes}
Keep this brief — it's meant to ground the abstraction of "proportion yes" in a single concrete moment a
participant experiences. This is also a good anchor to come back to later when we talk about each trial as
a binomial (yes/no) event.
:::

---

## A concrete example: a psychometric function

Classic psychophysics setup: on each trial, a stimulus of a given intensity is presented, and the
participant says whether they detected it ("yes"/"no"). We repeat this many times at each of several
intensities — the same intensities shown on the previous slides.

```{r}
intensity <- 1:9
n_trials <- rep(40, 9)
n_yes <- c(0, 4, 8, 16, 25, 28, 37, 39, 40)
prop_yes <- n_yes / n_trials

data <- data.frame(intensity, n_yes, n_trials, prop_yes)
data
```

::: {.notes}
This is presented as if it were a real dataset — no generating code, no "true" parameters revealed. The
tutorial will later show you how a dataset like this could be simulated with known parameters, once you're
comfortable with the fitting tools themselves.
:::

---

## Plotting the data

```{r}
plot(data$intensity, data$prop_yes, pch = 19,
     xlab = "Stimulus intensity",
     ylab = "Proportion 'yes' responses",
     ylim = c(0, 1), main = "Psychometric function")
```

- The curve rises from near 0 to near 1 as intensity increases — an S-shape
- To describe this curve with a model, we need just two numbers

::: {.notes}
Point at the plot: roughly where does it start rising steeply, and roughly how gradual or sharp is the rise.
Don't name the parameters yet — that's the next slide.
:::

---

## The model

$$p(x) = 1 - e^{-(x/\alpha)^{\beta}}, \quad x \geq 0$$

This is the **Weibull cumulative distribution function** — see
[Wikipedia: Weibull distribution](https://en.wikipedia.org/wiki/Weibull_distribution) for the full
mathematical background.

- `alpha` (scale): a multiplicative scale parameter — rescales the intensity axis, which moves where the
  curve reaches about 63% *and* proportionally stretches or compresses how wide the transition is
- `beta` (shape): controls how sharply the curve rises, independent of where it sits

```{r}
psychometric_model <- function(alpha, beta, x) {
  y <- rep(0, length(x))
  valid <- x >= 0
  y[valid] <- 1 - exp(-(x[valid] / alpha)^beta)
  y
}
```

Let's see whether a plausible guess looks like our data:

```{r}
plot(data$intensity, data$prop_yes, pch = 19, ylim = c(0, 1),
     xlab = "Intensity", ylab = "P('yes')")
curve(psychometric_model(5, 3, x), add = TRUE, col = "blue")
```

::: {.notes}
Unlike the cumulative normal, the Weibull CDF has no dependency on a built-in distribution function at all —
it's written directly from its own formula, using only exp() and arithmetic. This is worth pointing out
explicitly: nothing here is a black box. The overlay is a nice payoff: alpha=5, beta=3 wasn't fit to this
data at all, it's just a reasonable guess, and it already tracks the general shape reasonably well, even
with the noise in a couple of the data points.
:::

---

## Why this shape? Weibull vs. alternatives

Several S-shaped curves could plausibly describe data like ours — cumulative normal, logistic, and Weibull
all rise smoothly from 0 to 1, and in practice any of them might fit reasonably well.

- Stimulus intensity can only ever be positive — there's no such thing as negative contrast or negative
  loudness
- The Weibull distribution is defined **only for x ≥ 0** by construction, which matches that constraint
  naturally
- Cumulative normal and logistic are technically defined across *all* real numbers, including negative x —
  in practice they can still fit the visible range fine, but Weibull's domain matches the physical
  measurement scale a little more naturally for this kind of data

Further reading: [Weibull distribution](https://en.wikipedia.org/wiki/Weibull_distribution),
[Normal distribution](https://en.wikipedia.org/wiki/Normal_distribution),
[Logistic distribution](https://en.wikipedia.org/wiki/Logistic_distribution)

- None of these is universally "correct" — the choice depends on what you're assuming about the underlying
  process and the measurement scale of your stimulus
- We'll return to this kind of question properly in Model Comparison

::: {.notes}
Keep this at the level of "different motivations, no universal winner" rather than a deep dive — the full
apparatus for adjudicating between model shapes rigorously is Weeks 6–7.
:::

---

## What does `alpha` do?

```{r}
curve(psychometric_model(3, 3, x), from = 0, to = 9, ylim = c(0, 1), col = "red",
      xlab = "Intensity", ylab = "P('yes')", main = "Varying alpha (beta fixed at 3)")
curve(psychometric_model(5, 3, x), add = TRUE, col = "blue")
curve(psychometric_model(7, 3, x), add = TRUE, col = "darkgreen")
legend("topleft", legend = c("alpha = 3", "alpha = 5", "alpha = 7"),
       col = c("red", "blue", "darkgreen"), lty = 1)
```

- `alpha` is a **scale** parameter, not a pure shift: increasing it moves the curve right *and* stretches its
  width — notice the three curves above aren't just shifted copies of each other, the alpha=7 curve is
  visibly more gradual than the alpha=3 curve
- A useful fixed point regardless of `beta`: the curve always reaches about **63%** at `x = alpha` (a
  property of the Weibull's mathematical form)
- Psychologically: how sensitive the person is overall — a lower `alpha` means they reach a given detection
  level at a fainter stimulus

::: {.notes}
Let students watch alpha and only alpha change; unlike a pure location parameter, the curves here should
visibly differ in width too, not just position — worth pausing on this since it's a genuine (and useful)
contrast with the cumulative normal's mu, which really was a pure shift. The 63% fact is worth stating
explicitly (it comes from 1 - e^-1 ≈ 0.632) — it's the Weibull's fixed reference point, playing a similar
role to "mu is the 50% point" for the cumulative normal, just without the "no width change" guarantee.
:::

---

## What does `beta` do?

```{r}
curve(psychometric_model(5, 0.7, x), from = 0, to = 9, ylim = c(0, 1), col = "red",
      xlab = "Intensity", ylab = "P('yes')", main = "Varying beta (alpha fixed at 5)")
curve(psychometric_model(5, 3, x), add = TRUE, col = "blue")
curve(psychometric_model(5, 8, x), add = TRUE, col = "darkgreen")
legend("topleft", legend = c("beta = 0.7", "beta = 3", "beta = 8"),
       col = c("red", "blue", "darkgreen"), lty = 1)
```

- `beta` controls **how sharply** the curve rises — all three curves pass through the same 63% point (at
  `x = 5`) but climb at very different rates
- Psychologically: how *reliably* the person's responses track the physical stimulus

---

## Why don't the dots sit exactly on the curve?

- The model gives a single, *deterministic* prediction for each intensity
- But real data always has some trial-to-trial noise
- We can think of the data as: **model prediction + random error**

$$\text{observed} = \text{model}(\alpha, \beta, x) + \text{error}$$

- This is *why* fitting can get close but never perfect

---

## How wrong is a guess? — RMSD

Following the textbook, we use the **root mean squared deviation (RMSD)** rather than a raw sum of squares:

$$\text{RMSD} = \sqrt{\frac{\sum_{j=1}^{J}(d_j - p_j)^2}{J}}$$

```{r}
rmsd <- function(params, data) {
  alpha <- params[1]
  beta <- params[2]
  predicted <- psychometric_model(alpha, beta, data$intensity)
  sqrt(mean((data$prop_yes - predicted)^2))
}

rmsd(c(3, 0.5), data)   # a bad-ish guess
rmsd(c(5, 3), data)     # a better guess
```

- Bigger number = worse fit; our job is to make it as small as possible
- RMSD is in the same units as the data (here, a proportion) and doesn't grow just because we collected more
  data points, unlike a raw sum of squares — this is why the textbook (and most of the field) prefers it

::: {.notes}
Run both lines live. The "same units as the data" point is worth emphasizing explicitly — students will meet
RMSD values as pure numbers all course, and knowing they're interpretable (e.g., "off by 3 percentage
points on average") makes them far less abstract.
:::

---

## Seeing RMSD: a few deliberately bad fits

```{r}
#| fig-width: 10
#| fig-height: 8
candidates <- list(c(alpha = 5, beta = 3), c(alpha = 5, beta = 0.3),
                    c(alpha = 2, beta = 3), c(alpha = 8, beta = 5))

par(mfrow = c(2, 2))
for (cand in candidates) {
  pred_curve <- psychometric_model(cand["alpha"], cand["beta"], seq(0, 9, length.out = 200))
  r <- round(rmsd(cand, data), 3)
  plot(data$intensity, data$prop_yes, pch = 19, ylim = c(0, 1),
       xlab = "Intensity", ylab = "P('yes')",
       main = paste0("alpha=", cand["alpha"], ", beta=", cand["beta"], "  (RMSD=", r, ")"))
  lines(seq(0, 9, length.out = 200), pred_curve, col = "blue")
}
par(mfrow = c(1, 1))
```

::: {.notes}
Deliberately pick one good fit and three clearly bad ones (wrong scale, wrong shape, both wrong) so students
can connect "this curve visually misses the dots" to "and that's why its RMSD number is high." Point out
explicitly which panel has the lowest RMSD and confirm it's also the one that looks best by eye.
:::

```{r}
#| echo: false
ojs_define(data_ojs = data)
```

---

## Try it yourself — live

```{ojs}
//| echo: false
viewof alpha_slider = Inputs.range([1, 9], {value: 5, step: 0.1, label: "alpha (scale)"})
viewof beta_slider = Inputs.range([0.1, 6], {value: 3, step: 0.05, label: "beta (shape)"})
```

```{ojs}
//| echo: false
function weibullCdf(x, alpha, beta) {
  return x >= 0 ? 1 - Math.exp(-Math.pow(x / alpha, beta)) : 0;
}

predicted_points = transpose(data_ojs).map(d => ({
  ...d,
  pred: weibullCdf(d.intensity, alpha_slider, beta_slider)
}))

curve_points = d3.range(0, 9.01, 0.05).map(x => ({
  intensity: x,
  pred: weibullCdf(x, alpha_slider, beta_slider)
}))

rmsd_value = Math.sqrt(
  d3.mean(predicted_points, d => (d.prop_yes - d.pred) ** 2)
)

Plot.plot({
  x: {domain: [0, 9], label: "Intensity"},
  y: {domain: [0, 1], label: "P('yes')"},
  marks: [
    Plot.line(curve_points, {x: "intensity", y: "pred", stroke: "steelblue"}),
    Plot.dot(transpose(data_ojs), {x: "intensity", y: "prop_yes", r: 5, fill: "black"})
  ],
  caption: `RMSD = ${rmsd_value.toFixed(3)}`
})
```

- Drag the sliders and watch RMSD change live — try to find the combination that minimizes it
- This is exactly what `optim()` will do automatically, in a moment

---

## Grid search: try everything

```{r}
alpha_vals <- seq(1, 9, length.out = 100)
beta_vals <- seq(0.1, 6, length.out = 100)
grid <- expand.grid(alpha = alpha_vals, beta = beta_vals)

grid$rmsd <- apply(grid, 1, function(row) rmsd(row, data))

dim(grid)
head(grid, 10)
```

```{r}
best <- grid[which.min(grid$rmsd), ]
best
```

- `dim(grid)` confirms we're testing 100 x 100 = 10,000 combinations — every row in `head(grid, 10)` above is
  one candidate `(alpha, beta)` pair, with its RMSD already computed

---

## Seeing the cost surface

```{r}
library(lattice)
cost_palette <- colorRampPalette(c("white", "steelblue", "midnightblue"))(100)
levelplot(rmsd ~ alpha * beta, data = grid,
          main = "RMSD across parameter combinations",
          xlab = "alpha (scale)", ylab = "beta (shape)",
          col.regions = cost_palette)
```

- Dark blue = low RMSD = good fit
- Our best-fitting combination sits at the bottom of that "valley"

---

## Grid search doesn't scale

| # parameters | grid points per parameter | total combinations to test |
|:---:|:---:|:---:|
| 2 (today's model) | 100 | 10,000 |
| 4 | 100 | 100,000,000 |
| 10 (a small neural network) | 100 | 10^20 |

- Two parameters: totally fine on a laptop
- A handful more: computationally impossible to brute-force
- We need something smarter that *searches intelligently* instead of trying everything

::: {.notes}
Read the last row out loud slowly so the scaling problem registers as absurd rather than abstract.
:::

---

## What is `optim()` actually doing?

Two physical metaphors the textbook uses:

- Imagine the error surface carved out of wood, and you drop a marble onto it — it rolls downhill and comes
  to rest at the bottom
- Or: a parachutist dropped onto unfamiliar terrain at night, who can't see the whole landscape, only feels
  which way is downhill from where she's standing right now, and takes small steps accordingly

- Start somewhere (your initial guess), move downhill, repeat until no nearby step improves things
- Much faster than grid search — it doesn't need to check every point on the map
- But: it can walk straight into a small valley that isn't the lowest one overall — a **local minimum**

**Live class demo:** [i-am.ai/gradient-descent.html](https://www.i-am.ai/gradient-descent.html) — an
interactive game where you nudge a point downhill on a landscape by hand, before we get to the specifics of
the exact method `optim()` uses. (Note: this game illustrates *gradient* descent, which uses the local slope
directly — `optim()`'s default method, covered next, doesn't use slope information at all, but the
"search downhill, watch out for false valleys" intuition transfers directly.)

---

## Meet the method: Nelder-Mead Simplex

When you call `optim()`, its default method is the **Nelder-Mead Simplex** algorithm (Nelder & Mead, 1965).

- A "simplex" is a simple geometric shape: for 2 parameters, it's a **triangle**, with each corner being one
  candidate `(alpha, beta)` combination to test — for example, one corner might be `(1, 2)`, another
  `(4, 5)`, another `(6, 6)`
- At each step, the corner with the *worst* fit is moved: either **reflected** to the opposite side of the
  triangle (and possibly stretched further, if that direction looks promising), or the whole triangle
  **contracts** inward if reflection doesn't help
- Repeating this, the triangle "tumbles" downhill and eventually shrinks to a single point — the best-fitting
  estimate

---

## Watching a simplex tumble downhill

```{r}
#| echo: false
triangle <- rbind(c(2, 1), c(3, 0.3), c(1.5, 3))
n_steps <- 7
triangle_history <- vector("list", n_steps)
triangle_history[[1]] <- triangle

for (step in 2:n_steps) {
  costs <- apply(triangle, 1, function(p) rmsd(p, data))
  worst <- which.max(costs)
  centroid <- colMeans(triangle[-worst, , drop = FALSE])
  reflected <- centroid + (centroid - triangle[worst, ])
  reflected_cost <- rmsd(reflected, data)
  if (reflected_cost < costs[worst]) {
    triangle[worst, ] <- reflected           # reflection improved things -- accept it
  } else {
    triangle[worst, ] <- centroid + 0.5 * (triangle[worst, ] - centroid)  # otherwise, contract instead
  }
  triangle_history[[step]] <- triangle
}

rmsd_matrix <- matrix(grid$rmsd, nrow = length(alpha_vals), ncol = length(beta_vals))
cost_palette <- colorRampPalette(c("white", "steelblue", "midnightblue"))(100)
image(alpha_vals, beta_vals, rmsd_matrix, col = cost_palette,
      xlab = "alpha (scale)", ylab = "beta (shape)",
      main = "A simplex tumbling downhill on the real RMSD landscape")
contour(alpha_vals, beta_vals, rmsd_matrix, add = TRUE, nlevels = 8)

step_colors <- colorRampPalette(c("black", "red"))(n_steps)
for (step in 1:n_steps) {
  tri <- triangle_history[[step]]
  polygon(tri[, 1], tri[, 2], border = step_colors[step], lwd = 2)
}
legend("topright", legend = c("Step 1 (black)", paste0("Step ", n_steps, " (red)")),
       col = c("black", "red"), lty = 1, lwd = 2, bg = "white")
```

::: {.notes}
This is the same RMSD landscape and color scale as the "Seeing the cost surface" slide, drawn with image() +
contour() instead of a lattice levelplot so we can overlay the triangle on top of it. The simplex here is a
genuinely simulated sequence, evaluated using our actual rmsd() function on the real data: each step either
reflects the worst corner through the centroid of the other two (if that improves things) or contracts it
halfway toward the centroid instead (if reflection would make things worse). That reflect-or-contract choice
is the fix that keeps the sequence well-behaved — earlier versions that always accepted the reflection could
overshoot and oscillate instead of cleanly converging. Between any two consecutive triangles shown, exactly
one corner moves and the other two stay fixed — that's worth pointing out explicitly if students ask why two
of the three points always line up between neighboring steps. It's still simplified relative to the full
algorithm (no expansion step) but should visibly converge toward the same dark valley from the "Seeing the
cost surface" slide, color-coded from black (step 1) to red (final step). The code itself isn't the point
here — focus on the figure.
:::

---

## `optim()`: automatic search

```{r}
fit_rmsd <- optim(
  par = c(alpha = 4, beta = 1),
  fn = rmsd,
  data = data
)
fit_rmsd$par
fit_rmsd$value
```

- `par`: our starting guess
- `fn`: the function to minimize
- `method`: which search algorithm to use — defaults to `"Nelder-Mead"`, but can be set explicitly to
  something else, e.g. `"BFGS"` (Broyden–Fletcher–Goldfarb–Shanno):

```r
optim(par = c(alpha = 4, beta = 1), fn = rmsd, data = data, method = "BFGS")
```

::: {.notes}
Run the first version live and compare to `best` from the grid search — they should be very close. The
method="BFGS" line is just shown, not run — the point is purely that method is a normal named argument you
can override, not a deep dive into what BFGS does differently.
:::

---

## Limitations: getting stuck in a local minimum

Not every landscape looks like our psychometric-function example. Here's a synthetic (illustrative, not
psychologically meaningful) example with a bumpier landscape:

```{r}
#| echo: false
toy_cost <- function(x, y) (x - 3)^2 + (y + 1)^2 + 15 * sin(x) * cos(y)

x_vals <- seq(-6, 10, length.out = 100)
y_vals <- seq(-6, 6, length.out = 100)
toy_grid <- expand.grid(x = x_vals, y = y_vals)
toy_grid$cost <- with(toy_grid, toy_cost(x, y))

levelplot(cost ~ x * y, data = toy_grid, main = "A landscape with multiple valleys")
```

- Several dark patches = several **local minima**; the single darkest patch overall is the **global
  minimum** — what we actually want
- There's no guarantee that what `optim()` finds is the global minimum rather than a local one — this is
  true of essentially all parameter estimation methods, not just Simplex
- One practical mitigation: run `optim()` several times from different starting values; if they all agree,
  that's good evidence you've found the global minimum

::: {.notes}
Keep the toy landscape framed explicitly as synthetic (made-up x,y, not our alpha,beta) so students don't
think our psychometric model secretly has hidden local minima it doesn't have.
:::

---

## Limitations: struggling with many parameters

- Simplex becomes inefficient and unreliable with roughly 4 or more parameters
- We'll meet methods much better suited to many-parameter models when we get to neural networks in a few
  weeks

::: {.notes}
A direct, honest hand-off to the neural network sessions, where gradient-based training methods that scale
to far more parameters take over.
:::

---

## A different kind of question

So far: "which parameters make my prediction closest to the data?"

A related but different question:

**"Which parameters make the data I actually observed most *probable*?"**

This is maximum likelihood estimation (MLE).

::: {.notes}
Slow down here — this is the conceptual pivot point of the lecture.
:::

---

## From simulating to evaluating

`rbinom()` **generates** simulated data from known parameters — you'll see this used in the tutorial to
build a dataset with known "true" values:

```{r}
rbinom(1, size = 40, prob = 0.8)  # parameters -> simulated data
```

`dbinom()` does the reverse — it **evaluates** how probable a specific outcome was, given parameters:

```{r}
dbinom(32, size = 40, prob = 0.8)  # data + parameters -> probability
```

- `rbinom`: parameters → data
- `dbinom`: data + parameters → probability (what MLE needs)
- Each "yes"/"no" trial in our psychometric experiment is exactly this kind of binomial event

::: {.notes}
The tutorial introduces rbinom() properly when it shows students how to simulate their own dataset with
known parameters, later in the practical — this slide just introduces dbinom() as its natural counterpart
for the lecture's purposes.
:::

---

## A simple case: detecting a fixed stimulus

Imagine presenting the same intensity-5 patch 10 times in a row, and recording "yes" (1) or "no" (0) each
time:

```{r}
detected <- c(1, 0, 1, 1, 0, 1, 1, 1, 0, 1)  # 10 presentations, 7 "yes"
sum(detected)
```

- If the true detection probability at this intensity is `p`, how likely is *this exact sequence* of 10
  responses?
- Each trial is a small binomial event (`size = 1`); `dbinom()` gives us exactly that probability

```{r}
dbinom(sum(detected), size = 10, prob = 0.5)
dbinom(sum(detected), size = 10, prob = 0.7)
```

::: {.notes}
Ask students to guess which probability they think is "more likely" before running the second line. Worth
noting that each intensity level in our psychometric data is the same kind of binomial count — just with
size = 40 instead of size = 1 per observation, and aggregated into a single count rather than shown trial by
trial.
:::

---

## The likelihood function

```{r}
p_vals <- seq(0.01, 0.99, by = 0.01)
likelihood <- sapply(p_vals, function(p) dbinom(sum(detected), 10, p))

plot(p_vals, likelihood, type = "l",
     xlab = "Candidate p", ylab = "Likelihood of observed data")
p_vals[which.max(likelihood)]
```

- The peak of this curve is the **maximum likelihood estimate**
- It matches the intuitive answer: 7/10 = 0.7

::: {.notes}
Point out that this plot has exactly the same shape/logic as the RMSD surface earlier, just flipped: we want
the peak here instead of the valley.
:::

---

## Seeing likelihood as heights, not distances

```{r}
cand_good <- c(5, 3)     # a good-fitting guess
cand_bad  <- c(2, 0.5)   # a clearly poor-fitting guess

lik_good <- dbinom(data$n_yes, data$n_trials,
                    psychometric_model(cand_good[1], cand_good[2], data$intensity))
lik_bad  <- dbinom(data$n_yes, data$n_trials,
                    psychometric_model(cand_bad[1], cand_bad[2], data$intensity))

barplot(rbind(lik_good, lik_bad), beside = TRUE, names.arg = data$intensity,
        col = c("darkgreen", "red"), xlab = "Intensity",
        ylab = "Probability of the observed count",
        legend.text = c("Good-fitting alpha,beta", "Poor-fitting alpha,beta"))
```

- RMSD pictures *distance* (smaller is better)
- Likelihood pictures *probability, one intensity level at a time* (taller is better)

::: {.notes}
The green bars (good params) should now visibly and clearly tower over the red bars (poor params) at most
intensity levels — the two candidate parameter sets were deliberately chosen to be quite different from each
other so the contrast reads clearly on the plot. This slide gives the *per-point* view — the next slide
zooms out to the *overall* view across all points at once, which is the number we actually want to maximize.
:::

---

## Seeing the *overall* likelihood

Fitting doesn't care about one data point at a time — it cares about the **joint** probability of the whole
dataset together. We get that by multiplying (or, in log terms, adding) the per-point values from the
previous slide:

```{r}
loglik_good <- sum(log(lik_good))
loglik_bad  <- sum(log(lik_bad))

barplot(c("Good-fitting alpha,beta" = loglik_good, "Poor-fitting alpha,beta" = loglik_bad),
        ylab = "Total log-likelihood",
        col = c("darkgreen", "red"))
```

- We add up *logged* probabilities rather than multiplying raw ones — raw probabilities are tiny numbers that
  quickly underflow to zero once you multiply many together, whereas logs turn that multiplication into a
  simple, well-behaved sum
- Whichever parameter set gives the **higher** total log-likelihood is the better overall fit — exactly
  mirroring "whichever parameter set gives the lower RMSD" from earlier

::: {.notes}
This closes the loop opened by the per-point bars slide: individual tall bars are nice, but what we actually
optimize is the sum across every data point at once. Emphasize the log point isn't just a mathematical
nicety — with real datasets of hundreds of trials, the raw joint probability becomes a number too small for
a computer to represent accurately, so working in logs is a practical necessity, not just convention.
:::

---

## Back to the psychometric function

- We don't have single yes/no trials at each intensity — we have counts (e.g., 28 out of 40 said "yes")
- `dbinom()` still works, just with `size = n_trials`
- We work with *log*-likelihood and *minimize* the negative log-likelihood (so we can keep using `optim()`,
  which minimizes by default)

```{r}
nll <- function(params, data) {
  alpha <- params[1]; beta <- params[2]
  predicted <- psychometric_model(alpha, beta, data$intensity)
  predicted <- pmin(pmax(predicted, 1e-6), 1 - 1e-6)
  -sum(dbinom(data$n_yes, data$n_trials, predicted, log = TRUE))
}
```

::: {.notes}
The pmin/pmax line keeps predicted probabilities inside (0,1) to avoid numerical errors — housekeeping, not
conceptually important, but flag it so students aren't confused in the tutorial when they see it in the
code.
:::

---

## Fitting by maximum likelihood

```{r}
fit_mle <- optim(par = c(alpha = 4, beta = 1), fn = nll, data = data)
fit_mle$par
fit_rmsd$par
```

- Two different logics (RMSD vs. likelihood), often very similar answers

::: {.notes}
This comparison is the payoff of the whole lecture. If time allows, plot both fitted curves over the data
points side by side — visually near-identical.
:::

---

## Summary

- A model has parameters, and each parameter has a psychological meaning worth stating in words — here,
  scale and shape of a Weibull curve, chosen because stimulus intensity can only be positive
- The choice of S-shape (Weibull, cumulative normal, logistic) reflects different assumptions and
  priorities, not a single "correct" answer — the logistic version, in particular, is the same shape used
  inside neural network units later in the course
- Data = model + error — that's why a fit is never perfect
- RMSD → grid search → `optim()`'s default Nelder-Mead simplex, which tumbles downhill but can get stuck in
  a local minimum on harder landscapes, and doesn't scale well beyond about 3–4 parameters — both
  limitations we'll revisit when methods better suited to complex, many-parameter landscapes come up in the
  neural network sessions
- Maximum likelihood asks a different question ("most probable data") but is solved the same way (minimize a
  function); `dbinom()` is the reverse of `rbinom()`, which you'll use in the tutorial to simulate your own
  dataset; the per-point view (bar heights) and the overall view (summed log-likelihood) are two zoom levels
  on the same idea
- Next: the tutorial — you'll do all of this yourselves, simulate your own dataset with known parameters to
  check recovery, and finally compare today's Weibull model against the linear model you already know

::: {.notes}
Close by pointing them to the tutorial handout. Flag that the tutorial's final exercise (linear vs. Weibull)
has an extra payoff: a straight line has no way to respect the 0–1 probability bounds that a real
psychometric function must, which gives students a very concrete, visual reason a model's *shape* matters
beyond just its fit quality — nice bridge into Model Comparison later in the course.
:::
